Design Document for Dimension Reduction in R Package glmm

### 2020-07-16 DesignDocs/DesignDocDimReduction.tex
started design doc for dim reduction

\title{Design Document for R Package glmm}

@@

\begin{abstract}
This design document describes the process of performing Monte Carlo maximum likelihood (MCML) for generalized linear mixed models.  First, penalized quasi-likelihood (PQL) estimates are calculated, which help generate simulated random effects. Then, the Monte Carlo likelihood approximation (MCLA)  is calculated using the simulated random effects. Next, the MCLA is maximized to find Monte Carlo maximum likelihood estimates, the corresponding Fisher Information, and other statistics. Additional inference is then possible, including confidence intervals for the parameters.
\end{abstract}

\section{Theory}

Let $y=(y_1, \ldots, y_n)'$ be a vector of observed data. Let $u=(u_1,\ldots,u_q)'$ be a vector of unobserved, normally-distributed random effects centered at 0 with variance matrix $D$. Let $\beta$ be a vector of $p$ fixed effect parameters and let $\nu$ be a vector of $T$ variance components for the random effects so that $D$ depends on $\nu$. Let $\theta=(\beta ,  \nu)'$ be a vector containing all unknown parameters. Then the data $y$ are distributed conditionally on the random effects according to $f_\theta(y|u)$ and the random effects are distributed according to $f_\theta(u)$. Although $f_\theta(u)$ does not actually depend on $\beta$ and $f_\theta(y|u)$ does not depend on $\nu$, we write both densities with $\theta$ to keep notation simple in future equations.

Since $u$ is unobservable, the log likelihood must be expressed by integrating out the random effects:
\begin{align}
l(\theta)=\log \int f_\theta(y|u) f_\theta(u) \; du
\end{align}
For most datasets, this integral is intractible. In these cases, performing even basic inference on the likelihood is not possible. Rather than evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood. Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.  MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is the mixture distribution specified in section \ref{sec:genRand}.

Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}. \label{eq:MCLAval}
\end{align}
When $\tilde{f}$ does not depend on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
\nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)       \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \nonumber\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)   - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  - \nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \label{eq:MCLAhessian}
\end{align}
Details for these derivatives can be found in section \ref{sec:calcsindep}. Calculation for the gradient and Hessian when $\tilde{f}$ is dependent on $\theta$ can be found in section \ref{sec:calcs}.

Now any likelihood-based inference, such as maximum likelihood, can be performed on $l_m(\theta)$ and its derivatives.

\section{Model fitting function} 
The model fitting function is be the primary function the user  uses. The user  specifies the response and the predictors using the R formula mini-language as interpreted by \texttt{model.matrix}. Let $n$ be the observed sample size and recall that $p$ is the number of fixed effects.  As a result of the user specifying the fixed effects, the model fitting function creates matrix $X$, which has dimensions $n \times p$.

The user  specifies the random effects in the same way as the fixed effects. This is also how users of  \texttt{reaster} (of R package
 \texttt{aster} \citep{aster-package}) specify the random effects. That is, random effects are expressed using the R formula mini-language. Let $q_t$ be the number of random effects associated with variance component $\nu_t$. When the random effects are specified, a list of $T$ model matrices are created. The $t$th model matrix $Z_t$ has dimensions $n \times q_t$.

The user also specifies the expontial family (details in section \ref{sec:fam}), the name of the data set, and the names of the variance components.

Thus, the following is a sample command with fixed predictors $x_1$ and $x_2$ and with random effects created by categorical variables $school$ and $classroom$:
\begin{verbatim}
glmm(y ~ x1+ x2, list(~0+school,~0+classroom),  family.glmm="bernoulli.glmm", 
data=schooldat,varcomps.names=c("school","classroom"),varcomps.equal=c(1,2),
debug=FALSE)
 \end{verbatim}

It is possible that the user could want some variance components to be set equal. For example, \pcite{coull:agresti:2000} influenza dataset contains four years with random effects for each year. The authors want the within-year variance components to be equal.  There are also variance components for subject-specific intercepts and for the decreased susceptibility to illness in year 4 (since the strain of flu during year 4 was a repeat of a previous year). Suppose \texttt{year} is a categorical variable with four levels, and \texttt{year1} through \texttt{year4} are dummy variables. Thus, the call  contains these arguments:
\begin{verbatim}
glmm(y~year,list(~0+subject,~0+year1,~0+year2,~0+repeat,~0+year3,~0+year4),
 varcomps.equal=c(1,2,2,3,2,2), varcomps.names=c("subject","year","repeat"),
data=flu,family.glmm=bernoulli.glmm)
\end{verbatim}

Initially, \texttt{model.matrix}  makes $Z$ into a list of 6 design matrices. Since we have 3 distinct variance components, we want 3 design matrices. We take the design matrices that share a variance component and use \texttt{cbind} to bind them together. The result is 3 model matrices in the $Z$ list, one for each variance component. We  want to put them in order (1,2,3 according to \texttt{varcomps.equal}) so that the names for each model matrix  (``subject'',``year'',``repeat'') are correctly assigned.  The variance estimates  are eventually listed in the model summary in this same order as well.

  After interpreting the model the user has specified, the next step is to find penalized quasi-likelihood (PQL) estimates.  The process of finding these estimates is detailed in section \ref{sec:pql}. The PQL estimates parameterize the importance sampling distribution $\tilde{f}(u_k)$ which generate the random effects.  More information on generating the random effects is in section \ref{sec:genRand}.

Next, \texttt{trust} is implemented to maximize  the MCLA objective function (details on the objective function are in section \ref{sec:objfun}). Finally, the function  returns parameter estimates, the log likelihood evaluated at those estimates, the gradient vector of the log likelihood,  the Hessian of the log likelihood at those estimates, information from the \texttt{trust} optimization, and other information.

%%Families%%
\section{Families} \label{sec:fam}

Let $g$ be the canonical link function and $\mu$ be a vector of length $n$ such that
\begin{align}
g(\mu) = X \beta + Z u
\end{align}
The choice of the link function is related to the distribution of the data, $ f_\theta(y|u)$. If the data have a  Bernoulli distribution, the link is $\logit(\mu)$. If the data have a Poisson distribution, the link is $\log (\mu)$. Currently, \texttt{glmm} offers only these two choices for the family, but any exponential family could work and can be easily added later.

 For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z u$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
\begin{align}
y' \eta - c(\eta) = \sum_i \left[ y_i \eta_i - c(\eta_i)  \right]
\end{align}

The user is required to specify the family in the model-fitting function. Once the family is specified, many family-specific functions are called. They are contained in an S3 class called ``glmm.family''. Each family function outputs a list including the family name (a character string such as ``bernoulli.glmm''), a function that calculates the value of the cumulant function $c(\eta)$,  a function that calculates the cumulant's first derivative $c'(\eta)$ with the derivative taken with respect to $\eta$, and  a function that calculates the cumulant's second derivative $c''(\eta)$.

The users provide the family in the model-fitting function by either enter the character string (``bernoulli.glmm''), the function (\texttt{bernoulli.glmm()}), or the result of invoking the function.  The following code for using the input to determine the family is adapted from \texttt{glm}.

\begin{verbatim}
logDensity<-function(family.glmm)
{
	if(is.character(family.glmm))
		family.glmm<-get(family.glmm,mode="function",envir=parent.frame())
	if(is.function(family.glmm))
		family.glmm<-family.glmm()
	if(!inherits(family.glmm,"glmm.family")) 
		stop(" 'family.glmm' not recognized") 
	return(family.glmm)
}
\end{verbatim}
We interpret this as follows.  If the user has entered the family as a string, go get the R object with that family name, either from the immediate environment or the parent environment.  If this has happened, \texttt{family.glmm} is now a function.  If \texttt{family.glmm} is a function (either because the user entered it as a function or because of the preceding step), invoke that function.  At this point, \texttt{family.glmm} should have  class ``\texttt{glmm.family}.'' If this is not the case (maybe because of a typo or maybe because they entered ``\texttt{poisson}'' rather than ``\texttt{poisson.glmm}''), then stop and return an error.

With this family of functions, calculating $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$ is as simple as: 
\begin{verbatim}
family.glmm$cum(args)
family.glmm$cp(args)
family.glmm$cpp(args)
\end{verbatim}

For the Bernoulli distribution, we calculate these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) as follows. In order to be careful with our computer arithmetic, we use the \texttt{log1p} function and the second set of equalities.  
\begin{align}
c(\eta_i) &= \log(1+e^{\eta_i}) =
  \begin{cases}
    \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
    \eta_i+\log(e^{-\eta_i}+1) & \text{if } \eta_i >0,
  \end{cases}\\
c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{1}{1+e^{-\eta_i}}\\
c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{1}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
\end{align}

 For Poisson, these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) are
\begin{align}
c(\eta_i)&=e^{\eta_i}\\
c'(\eta_i)&=e^{{\eta_{i}}}\\
c''(\eta_i)&= e^{{\eta_{i}}}.
\end{align}

Then we use these pieces to create the scalar $c(\eta)$, the vector $c'(\eta)$ and the matrix $c''(\eta)$. We calculate

\begin{align}
c(\eta)= \sum_i c(\eta_i).
\end{align}
 The vector $c'(\eta)$ has components $c'(\eta_i)$. The matrix $c''(\eta)$ is diagonal with diagonal elements $c''(\eta_i)$.

In the R function \texttt{glm}, the user can choose the link. The canonical link must be used in the \texttt{glmm} package so that we have an exponential family. The canonical link is included in \texttt{family.glmm}  in case the user does not know it already.

Also, this family of functions contains a check to ensure the data are valid given the family type. If \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should contain only $0$ and $1$. If \texttt{family.glmm} is \texttt{poisson.glmm}, then the data should be nonnegative.  If the data set does not pass the check, the check returns an error message.

\subsection{Redone in C}
Redone in C, I  have  separate functions for calculating $c(\eta)$, $c'(\eta)$, $c''(\eta)$: \texttt{cum3.c}, \texttt{cp3.c}, and \texttt{cpp3.c}. The inputs for each function are identical:  \texttt{eta} (an array of doubles), $\texttt{neta}$ (the length of \texttt{eta}), the type (to denote the family: $1$ indicates Bernoulli and $2$ indicates Poisson), and an array of doubles to contain the result. These are passed in as pointers. The functions  calculate  the cumulant function or one of its first two derivatives. Each function  contains a switch statement for the glmm family. The calculations for each of these functions have  been shown earlier in this section. This function is type void: rather than returning the cumulant or its derivatives, the pointers are changed to contain the results. This function is invoked by \texttt{el}, described in section \ref{sec:el}.

\section{Log density of the data (\texttt{el})}\label{sec:el}
This section provides details for the log density of the data and two of its derivatives.

\subsection{Equations}
Recall the log of the data density is
\begin{align}
\log f_\theta(y|u) &= y' \eta -c(\eta) \\
&= \sum_{i} y_{i} {\eta_{i}} - c({\eta_{i}})
\end{align}
where
\begin{align}
\eta=X\beta+Zu.
\end{align}
The derivative of this with respect to one component, $\eta_j$, is
\begin{align}
\dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u)  = y_j-c'(\eta_j).
\end{align}
The derivative of the component $\eta_j$ with respect to one of the fixed effect predictors, $\beta_{l}$, is
\begin{align}
\dfrac{\partial \eta_j}{\partial \beta_{l}} = X_{j{l}}
\end{align}

We'd like the derivative of the log of the data density with respect to $\beta$. The first step is using the chain rule as follows:
\begin{align}
\dfrac{\partial}{\partial \beta_{l}}  \log f_\theta(y|u) &= \dfrac{\partial \eta_j}{\partial \beta_{l}} \dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u) \\
&= \left[ y_j-c'(\eta_j) \right]  X_{j{l}}
\end{align}

The mixed partial derivative (with respect to $\beta_{l_1}$ and $\beta_{l_2}$) of the log data density can be written similarly:
\begin{align}
\dfrac{\partial^2}{\partial \beta_{l_1} \partial \beta_{l_2}}  \log f_\theta(y|u) &=\dfrac{\partial}{\partial \beta_{l_2}} \left( \left[ y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
&= \dfrac{\partial \eta_j}{\partial \beta_{l_2}} \dfrac{\partial}{\partial \eta_j} \left( \left[ y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
&= -X_{j{l_1}} X_{j{l_2}} c''(\eta_j) 
 \end{align}

Letting $c'(\eta)$ be a vector with components $c'(\eta_j)$, the first derivative of the log data density can be written in matrix form as:
\begin{align}
\dfrac{\partial}{\partial \beta}  \log f_\theta(y|u) = X' \left[ y-  c'(\eta)  \right].
\end{align}

Letting $ c''(\eta)$ be a diagonal matrix with components $c''(\eta_j)$, the second derivative of the log data density can be written in matrix form as:
\begin{align}
   \frac{\partial^2}{\partial \beta^2} \log f_\theta(y|u) =   X' [ -  c''(\eta) ] X
\end{align}

\subsection{Redone in C}
The C function to calculate the value of the log data density and its two derivatives are called by reference. The following pointers are passed in: double \texttt{Y}, double \texttt{X}, int \texttt{nrowX}, int \texttt{ncolX}, double \texttt{eta}, int \texttt{family}, double \texttt{elval}, double \texttt{elgradient}, double \texttt{elhessian}. The pointers  \texttt{elval}, \texttt{elgradient} and \texttt{elhessian} are zeros before \texttt{el.C} is invoked. Invoking \texttt{el.C}  then places the calculatedvalue of the log data density and two derivatives into \texttt{elval}, \texttt{elgradient} and \texttt{elhessian}.\\

The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, \texttt{cpp3.C} to calculate the Hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.

\section{Random effect generation and calculations}
This section is focused on a vector of random effects $u$ with length $q$.  Section \ref{sec:getEk} details the  relationship between  $\nu$ and $D$.
Section \ref{sec:ftwiddle} states the importance sampling distribution.
 Section \ref{sec:genRand} describes the process of generating the random effects. Section \ref{sec:distRand} explains how to evaluate the distribution of the random effects and its first two derivatives.

%%getEk
\subsection{Constructing $D$ (\texttt{getEk})}\label{sec:getEk}
Recall that, for this version of the package, we assume that $D$ is diagonal.  Let $\nu$ be length $T$ with components $\nu_t$, and let $E_t,t=1,\ldots,T$ be diagonal matrices with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$. That is, the diagonal entries  of $E_t$ indicate whether that random effect has $\nu_t$ as its variance component. Then $D=\sum_{t=1}^T\nu_t  E_t $.

Recall $q_t$ is the number of random effects associated with variance component $\nu_t$. Then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.  
The model fitting function of \texttt{glmm} has made  \texttt{z} into a list with $T$ design matrices (one for each distinct variance component).  In order to create $E_t$, we need to  count the  number of columns ($q_t$) for each design matrix in the list.
Since $D$  is $q \times q$, each $E_t$  is $q \times q$.   $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal. We continue this process to complete the construction of $E_t, t=1,\ldots, T$.

\subsection{Importance sampling distribution $\tilde{f}(u_k)$}\label{sec:ftwiddle}
 Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale. That is $s$ is defined such that, $u= D^{1/2}s$. Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t},t=1,\ldots,T$.  Prior to generating random effects, the model fitting function of \texttt{glmm} has performed PQL and recorded the PQL estimates for $\beta$, $\sigma$ and  $s$. The PQL estimates are denoted by $\beta^*$, $s^*$ and $\sigma^*$. Let
\begin{align}
A^*=\sum_{t=1}^T E_t \sigma^*_t
\end{align}\
and
\begin{align}
 D^* =A^*A^*
\end{align}
be matrices based on PQL estimates. We can ``unstandardize'' our PQL-based random effects:
\begin{align}
u^*&=A^*s^*.
\end{align}

Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$.  Let $f(u|\mu,\Sigma)$ denote the pdf for $u \sim N(\mu,\Sigma)$.  Let $\grave{f}(u|0,D^*)$ denote the pdf for $u \sim t_\zeta$, a $q$-dimensional multivariate  $t$ with mean $0$, scale matrix $D^*$, and $\zeta$ degrees of freedom. Then the importance sampling distribution is:
\begin{align}
 \tilde{f}(u) = p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). \label{eq:ftwiddle}
\end{align}

The first component of the mixture distribution $t_\zeta$ is chosen to ensure  the gradient of the MCLA has a central limit theorem, which is proven in  section \ref{sec:CLT}.  The second component is chosen because it is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance. The last component is centered at the PQL guess $u^*$ and has a variance based on the Hessian of the PQL penalized likelihood.  The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y,u)$.

\subsection{Generating random effects (genRand)}\label{sec:genRand}

 Recall that $m$ is the overall Monte Carlo sample size.  Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$. We randomly determine $m_1,m_2,m_3$ according to the proportions $p_1,p_2,p_3$. More specifically, we generate $m$ random numbers from a distribution that is uniform between $0$ and $1$. Then $m_1$ is the frequency with which the randomly generated numbers are between $0$ and $p_1$,  $m_2$ is the frequency with which the randomly generated numbers are between $p_1$ and $p_1+p_2$, and $m_3=m-p_1-p_2$.
%Let $D$ and $A=D^{1/2}$ be determined by the current iteration of \texttt{trust}.

Once $m_1, m_2$ and $m_3$ are determined, we draw 
\begin{align}
u_k &\sim t_\zeta ,k=1,\ldots,m_1\\
u_k &\sim N(u \, | \, u^*, D^*) ,k=m_1+1,\ldots,m_1+m_2\\
u_k &\sim N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) ,k=m_1+m_2+1,\ldots,m.
\end{align}
Details for drawing from a nonstandard normal are in section \ref{sec:gennonstand}. We use R package \texttt{mvtnorm} to draw from a $t$ distribution.

\subsubsection{Finding the square root of a variance matrix}\label{sec:sqrtmat}
In order to generate draws from a nonstandard normal distribution, we need to calculate the square root of a variance matrix. When the variance matrix is diagonal (for example, $D^*$), we simply take the root of the diagonal elements.  When the variance matrix is not diagonal (for example, in the second component of the mixture distribution), we can use eigendecomposition. Eigendecompositions take a little bit more time than Cholesky decompositions but are more stable.

Let $\Sigma$ denote the nondiagonal variance matrix. Eigendecomposition of $\Sigma$ provides orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ (with diagonal entries $\lambda$ being the eigenvalues of $\Sigma$).  Then
\begin{align}
\Sigma^{1/2}= O \Lambda^{1/2} O'.
\end{align}
Finding $\Lambda^{1/2}$ is as easy as taking the root of the diagonal entries.  We know that $\Sigma^{1/2}$ is correct because
\begin{align}
\Sigma^{1/2} \Sigma^{1/2} &= O \Lambda^{1/2} O'O \Lambda^{1/2} O'\\ \nonumber
&= O \Lambda^{1/2}\Lambda^{1/2} O'\\\nonumber
&=O \Lambda O' \\\nonumber
&= \Sigma.
\end{align}

\subsubsection{Generating from nonstandard normal distributions}\label{sec:gennonstand}
 Construct $A^*=(D^*)^{1/2}$ and $((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} $ as described in section \ref{sec:sqrtmat}.

Let $\breve{u}_k,k=m_1+1\ldots,m$ be vectors of random effects that are drawn from the standard normal distribution. Then we center and scale $\breve{u}_k$ as follows:
\begin{align}
u_k&= u^*+A^* \, \breve{u}_k, k=m_1+1,\ldots,m_1+m_2\\
u_k&= u^*+ ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} \,  \breve{u}_k, k=m_1+m_2+1,\ldots,m.
\end{align}
 The result or centering and scaling is that
\begin{align}
 &u_k \sim N(u^*,D^*), \, k=m_1+1,\ldots,m_1+m_2\\
& u_k \sim N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}), \, k=m_1+m2+1,\ldots,m .
\end{align}

% 
%\subsubsection{NEED TO PUT THIS SOMEWHERE}
%
%
%Recall that $D$ is a diagonal with dimension $q$. $D$ has $q_1$ entries being $\nu_1$, $q_2$ entries being $\nu_2$, ..., and $q_T$ entries being $\nu_T$. Since the distribution of $u_k=A \breve{u}_k$ is $N(0,D)$, we write its pdf as:
%\begin{align}
%f(u_k|\nu)=f(A\breve{u}_k|\nu)&= (2 \pi)^{q/2} |D|^{-1/2} \exp(-0.5 (A \breve{u}_k)' D^{-1} (A \breve{u}_k))\\
%&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' A D^{-1} A \breve{u}_k)\\
%&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' \breve{u}_k)
%\end{align}
%So that 
%\begin{align}
%\log f(A \breve{u}_k|\nu) &\propto - \log |A| -\dfrac{1}{2} \breve{u}_k' \breve{u}_k\\
%&\propto -\dfrac{q_1}{2} \log \nu_1 +\ldots + -\dfrac{q_T}{2} \log \nu_T -\dfrac{1}{2} \breve{u}_k' \breve{u}_k
%\end{align}
%Then 
%\begin{align}
%\dfrac{\partial}{\partial \nu_t} \log f(A \breve{u}_k|\nu) = \dfrac{-q_t}{2 \nu_t}\\
%\dfrac{\partial^2}{\partial \nu_t^2} \log f(A \breve{u}_k|\nu) = \dfrac{q_t}{2 \nu_t^2}
%\end{align}
%
%Because only the first component of the mixture distribution contains any parameters,
%\begin{align}
%\dfrac{\partial}{\partial \nu} \log \tilde{f}(u) = \dfrac{p_1  \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})}.
%\end{align}
%and 
%\begin{align}
%\nabla  \log \tilde{f}(u) = \left(0   \; \; \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u) \right)
%\end{align}
%These are used in equations \ref{eq:MCLAgr} and \ref{eq:MCLAh}. The second derivatives, which are needed in equation \ref{eq:MCLAh}, are
%\begin{align}
%\dfrac{\partial^2}{\partial \nu^2} \log \tilde{f}(u) &= \dfrac{p_1  \dfrac{\partial^2}{\partial \nu^2} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} \\ \nonumber
%&- \dfrac{p_1^2 \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right) \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right)'}{\left[ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})\right]^2}.
%\end{align}
%and
%\begin{align}
%\nabla^2 \log \tilde{f}(u_k) = \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log \tilde{f} (u_k)  \end{bmatrix} 
%\end{align}
%
%

%%
%%This section is useful for calculating $\tilde{f}(u_k)$.  This section also provides two derivatives in some cases, which is useful for evaluating $f_\theta(u_k)$ and two derivatives. These are needed for the calculation in \eqref{eq:MCLA}.
%%
%%Section \ref{sec:tee} contains details that are not currently implemented, but could be. In this version, all random effects are assumed to be normally distributed and all simulated random effects are generated from a normal distribution.

\subsection{Evaluating the distribution of random effects (\texttt{distRand})}\label{sec:distRand}
This section discusses evaluating the distribution of the random effects, the gradient of the random effects, and the Hessian of the random effects. That is, the equations in this section provide  $\log f_\theta(u)$, $\nabla \log f_\theta(u)$, and $\nabla^2 \log f_\theta(u)$  for equation \ref{eq:MCLA}.

 We can  express the log unnormalized pdf of $N(0, D)$  as:
\begin{align}
\log f_\theta(u)= \log f (u| 0, D) = - q/2 \log (2 \pi) -1/2 \log |D| - (1/2) u' D^{-1} u
\end{align}

 To find derivatives of this, we are going to use the diagonal form of $D$. Recall that for every $\nu_t$, we can construct a matrix $E_t$ (with dimensions the same as matrix $D$) that has 1s on the diagonal elements corresponding to the elements of D that contain $\nu_t$ and 0s  elsewhere.

We can partition the random effects according to their variance components: $u=(U_1',...,U_T')'$.  Let $D_t$ be the variance matrix for $U_t$. $D_t$ has $q_t$ rows  and $q_t$ columns. Thus $D$ can be expressed as:
\begin{align}
D = \begin{bmatrix} D_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & D_T \end{bmatrix}
\end{align}

Since $D$ is diagonal, it follows that $D^{-1}$ is also diagonal with diagonal entries $\dfrac{1}{\nu_1}$,...,$\dfrac{1}{\nu_T}$.  Also, the assumption that $D$ is diagonal makes calculating the determinant of $D$  easy:
\begin{align}
|D|= \nu_1^{q_1}...\nu_T^{q_T}
\end{align}

 Taking these two pieces of information into account allows us to write the log density for $U$ as follows:
\begin{align}
\log f_\theta(u) &= -\dfrac{1}{2} \log |D| -\dfrac{1}{2} u' D^{-1} u\\
&= -\dfrac{1}{2} \left[  \sum_{t=1}^T q_t \log \nu_t   \right]  -\dfrac{1}{2} \sum_{t=1}^T \left[ \dfrac{1}{\nu_t} U_t'U_t   \right]
\end{align}

The first and second derivatives of each summand with respect to its associated $\nu_t$ are:
\begin{align}
\dfrac{\partial}{\partial \nu_t} \log f_\theta(u_t) = - \dfrac{q_t}{2 \nu_t} + \dfrac{1}{2 \nu_t^2}U_t'U_t
\end{align}
and 
\begin{align}
\dfrac{\partial^2}{\partial \nu_t^2} \log f_\theta(u_t) = \dfrac{q_t}{2 \nu_t^2}- \dfrac{1}{\nu_t^3} U_t'U_t.
\end{align}
Any other derivative is equal to 0. That is, for all $t_1 \neq t_2$,
\begin{align}
\dfrac{\partial}{\partial \nu_{t_1}} \log f_\theta(u_{t_2}) = 0.
\end{align}
Also,
\begin{align}
\dfrac{\partial}{\partial \beta} \log f_\theta(u_{t_2}) = 0.
\end{align}
Thus, if $\nu = (\nu_1,...,\nu_T)$, the gradient of the random effects distribution is the following vector of length $T$:
\begin{align}
\dfrac{\partial}{\partial \nu}  \log f_\theta(u) = \begin{bmatrix} - \dfrac{q_1}{2 \nu_1} + \dfrac{1}{2 \nu_1^2} U_1'U_1 & ... & - \dfrac{q_T}{2 \nu_T} + \dfrac{1}{2 \nu_T^2} U_T'U_T   \end{bmatrix} 
\end{align}

>>>>

\title{Design Document for Dimension Reduction in R Package glmm}

@@

\begin{abstract}
This design document describes the process of regularizing the random effects. For example, Sally and Tammy can share a random effect (and therefore be treated as the same salamander) if they have very similar random effects (such as $1.22$ and $1.24$). The goal of reducing the number of random effects is to reduce the computation time for MCLA.
\end{abstract}

\section{Background}

Let $y=(y_1, \ldots, y_n)'$ be a vector of observed data. Let $u$ be a $q$-vector of unobserved, normally-distributed random effects centered at 0 with variance matrix $D$. Let $\beta$ be a vector of $p$ fixed effect parameters and let $\nu$ be a vector of $T$ variance components for the random effects so that $D$ depends on $\nu$. Let $\theta=(\beta ,  \nu)'$ be a vector containing all unknown parameters. Then the data $y$ are distributed conditionally on the random effects according to $f_\theta(y|u)$ and the random effects are distributed according to $f_\theta(u)$. Although $f_\theta(u)$ does not actually depend on $\beta$ and $f_\theta(y|u)$ does not depend on $\nu$, we write both densities with $\theta$ to keep notation simple in future equations.

Since $u$ is unobservable, the log likelihood must be expressed by integrating out the random effects:
\begin{align}
l(\theta)=\log \int f_\theta(y|u) f_\theta(u) \; du
\end{align}
For most datasets, this integral is intractible. In these cases, performing even basic inference on the likelihood is not possible. Rather than evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood. Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.  .

Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}. \label{eq:MCLAval}
\end{align}

If $u_k$ were of length $q_2 < q$, this calculation would be less expensive. We therefore are trying to shorten the random effect vector $u$.

\section{Overview}

The package, in its current state has two main steps:
\begin{enumerate}
\item Use PQL to estimate the parameters and predict the random effects. (These predictions are then used to create the importance sampling distribution.)
\item Use MCLA to approximate the likelihood and conduct MCML.
  \end{enumerate}

I am figuring out one of these two options:
\begin{itemize}
\item[Plan A:] Change PQL so that similar random effects are collapsed into a shared, single random effect.
\item[Plan B:] Keep the PQL optimization as it is, and add another step in between the two steps. At this time, the similar random effects are collapsed into a shared, single random effect.
  \end{itemize}

  I'm not sure, but it seems like Plan A is better than Plan B. It makes sense that we'd want to try collapsing the random effects while making sure this simplification not ruin the fit of the model.

   In the currently used version of PQL, there is an optimization within an optimization. The inner optimization estimates the fixed effects and predicts the random effects. The outer optimization works on the variance components. \textbf{I believe that adding a penalty term to the inner loop will collapse the random effects.} We want the random effects to be similar to one other, so the distance (or the square of the distance, if we prefer a ridge-like penalty) between the random effects must be controlled.
